Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-06-14 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: WIP on description of OpenACC work

As the PSy layer iterates over columns, the single-core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer can
therefore be tailored for a particular hardware (such as multi-core,
many-core, GPUs, or some combination thereof) and software (such as
compiler, operating system, MPI library, \textit{etc}.) configuration
with no change to the Algorithm or Kernel layer code. This approach
therefore offers the potential for portable performance. In this work
we apply optimisations to the PSy layer manually. The development of a
tool to automate this process will be the subject of a future paper.

@@

In the Algorithm layer, fields (and grids) are treated as logically
global objects. Therefore, as part of creating the {PS}y{KA}l version
of NEMOLite2D, we represent fields with derived types instead of
arrays in this layer. These types then hold information about the
associated mesh and the extents of 'internal' and 'halo' regions as
well as the data arrays themselves. This frees the natural scientist
from having to consider these issues and allows for a certain degree
of flexibility in the actual implementation (e.g. padding for
alignment or increasing array extent to allow for other
optimisations). The support for this is implemented as a library
(which we term the GOcean Infrastructure) and is common to the
{PS}y{KA}l versions of both NEMOLite2D and Shallow.

In order to maximise the flexibility (and thus potential for
optimisation) of the {PS}y{KA}l version of NEMOLite2D, we made the
kernels as fine-grained as possible. This resulted in eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop (over the two dimensions of
the model grid) around each kernel call.

@@

We first describe the code transformations performed for the serial
version of NEMOLite2D.  We then move on to the construction of
parallel versions of the code using OpenMP and OpenACC. Again, we
describe the key steps we have taken in this process in order to
maximise the performance of the code. In both cases our aim is to
identify those transformations which must be supported by any tool
which seeks to auto-generate a performant PSy layer.

@@

The Time-update section includes several array copies where fields for
the current time-step become the fields at the previous time-step.
Initially we implemented these copies as 'built-in' kernels (in the
GOcean library) as they are specified in the algorithm layer. However,
we obtained better performance by simply in-lining these array copies
into the PSy layer.

@@

For this part of the work we began with the optimal {PS}y{KA}l version of
the code, as obtained after applying the various transformations
described in the previous section. As with the transformations of the
serial code, our purpose here is to determine the functionality
required of a tool that seeks to generate the PSy layer.

@@

Since data arrays are managed within the GOcean Library this
optimisation can again be implemented without changing the
natural-science code (\textit{i.e.} the Application and Kernel layers).

@@

@@

We first consider the performance of the code in serial and examine
the effects of the transformations described in
Section~\ref{sec_methodology}. Once we have arrived at an optimal form
for the serial version of NEMOLite2D we then investigate its parallel
performance on both CPU- and GPU-based systems.

@@

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the {PS}y{KA}l
apprach. He also experimented with using CUDA directly within the
original form of NEMOLite2D in order to quantify any performance
penalty incurred by taking the {PS}y{KA}l/OpenACC approach.

>>>>

As the PSy layer iterates over columns, the single-core performance
can be optimised by applying transformations such as manipulation of
loop bounds (e.g. padding for SIMD) and kernel
in-lining. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer can
therefore be tailored for a particular hardware (such as multi-core,
many-core, GPUs, or some combination thereof) and software (such as
compiler, operating system, MPI library, \textit{etc}.) configuration
with no change to the Algorithm or Kernel layer code. This approach
therefore offers the potential for portable performance. In this work
we apply optimisations to the PSy layer manually. The development of a
tool to automate this process will be the subject of a future paper.

@@

In the Algorithm layer, fields (and grids) are treated as logically
global objects. Therefore, as part of creating the {PS}y{KA}l version
of NEMOLite2D, we represent fields with derived types instead of
arrays in this layer. These types then hold information about the
associated mesh and the extents of `internal' and `halo' regions as
well as the data arrays themselves. This frees the natural scientist
from having to consider these issues and allows for a certain degree
of flexibility in the actual implementation (e.g. padding for
alignment or increasing array extent to allow for other
optimisations). The support for this is implemented as a library
(which we term the GOcean Infrastructure) and is common to the
{PS}y{KA}l versions of both NEMOLite2D and Shallow.

In re-structuring NEMOLite2D to conform to the {PS}y{KA}l separation
of concerns we must break up the computation into multiple
kernels. The more of these there are, the greater the potential for
optimisation of the {PS}y layer. This re-structuring gave eight distinct
kernels, each of which updates a single field at a single point (since
we have chosen to use point-wise kernels). With a little bit of
tidying/re-structuring, we found it was possible to express the
contents of the main time-stepping loop as a single invoke (a call to
the PSy layer) and a call to the I/O system
(Figure~\ref{FIG_psykal_nemolite2d_structure}). The single invoke
gives us a single PSy-layer routine which consists of applying each of
the kernels to all of the points requiring an update on the model
mesh. In its basic, unoptimised (`vanilla') form, this PSy-layer
routine then contains a doubly-nested loop (over the two dimensions of
the model grid) around each kernel call.

@@

We first describe the code transformations performed for the serial
version of NEMOLite2D.  We then move on to the construction of
parallel versions of the code using OpenMP and OpenACC. Again, we
describe the key steps we have taken in this process in order to
maximise the performance of the code. In both cases our aim is to
identify those transformations which must be supported by a tool
which seeks to auto-generate a performant PSy layer.

@@

The Time-update section includes several array copies where fields for
the current time-step become the fields at the previous time-step.
Initially we implemented these copies as 'built-in' kernels (in the
GOcean infrastructure) as they are specified in the algorithm
layer. However, we obtained better performance by simply in-lining
these array copies into the PSy layer.

@@

For this part of the work we began with the optimised {PS}y{KA}l
version of the code, as obtained after applying the various
transformations described in the previous section. As with the
transformations of the serial code, our purpose here is to determine
the functionality required of a tool that seeks to generate the PSy
layer.

@@

Since data arrays are managed within the GOcean infrastructure this
optimisation can again be implemented without changing the
natural-science code (\textit{i.e.} the Application and Kernel layers).

@@

\subsection{Construction of OpenACC-Parallel NEMOLite2D}
\label{sec_oacc_steps}

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the {PS}y{KA}l
approach.

\subsubsection{Data Movement}

Although the advent of technologies such as NVLink are alleviating the
bottleneck presented by the connection of the GPU to the CPU, it
remains critical to minimise data movement between the memory spaces
of the two processing units. In NEMOLite2D this is achieved by
performing all computation on the GPU. The whole time-stepping loop
can then be enclosed inside a single OpenACC \texttt{data} region and it
is only necessary to bring data back to the CPU for the purposes
of I/O.

\subsubsection{Kernel Acceleration}

Moving the kernels to execute on the GPU was achieved by using the
OpenACC \texttt{kernels} directive.

He also experimented with using CUDA directly within the
original form of NEMOLite2D in order to quantify any performance
penalty incurred by taking the {PS}y{KA}l/OpenACC approach.

@@

We first consider the performance of the code in serial and examine
the effects of the transformations described in
Section~\ref{sec_methodology}. Once we have arrived at an optimised
form for the serial version of NEMOLite2D we then investigate its
parallel performance on both CPU- and GPU-based systems.

@@

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the {PS}y{KA}l
apprach. He also experimented with using CUDA directly within the
original form of NEMOLite2D in order to quantify any performance
penalty incurred by taking the {PS}y{KA}l/OpenACC approach.
